---
title: "Class 11: AlphaFold"
author: "Mikaela Joya (PID: A17295169)"
format: pdf
---

```{r}
library(bio3d)

pdb <- read.pdb("hivpr_dimer_23119/hivpr_dimer_23119_unrelaxed_rank_002_alphafold2_multimer_v3_model_1_seed_000.pdb")

pdb
```

Make a vector of input PDB file names that we can read into R

```{r}
list.files("hivpr_dimer_23119/", 
           pattern = ".pdb",
           full.names = TRUE)
```

## Custom analysis of resulting models

For tidiness we can move our AlphaFold results directory into our RStudio project directory:

```{r}
results_dir <- "hivpr_dimer_23119/"
```

```{r}
pdb_files <- list.files(
  path = results_dir,
  pattern = "*.pdb",
  full.names = TRUE
)

basename(pdb_files)
```

```{r}
library(bio3d)

pdbs <- pdbaln(pdb_files, fit=TRUE, exefile="msa")
```

```{r}
pdbs
```

RMSD is a standard measure of structural distance between coordinate sets. We can use the `rmsd()` function to calculate the RMSD between all pairs models.

```{r}
rd <- rmsd(pdbs, fit=T)
```

```{r}
range(rd)
```

Draw a heatmap of these RMSD matrix values

```{r}
library(pheatmap)

colnames(rd) <- paste0("m",1:5)
rownames(rd) <- paste0("m",1:5)
pheatmap(rd)
```

Now lets plot the pLDDT values across all models.

```{r}
pdb <- read.pdb("1hsg")
```

```{r}
plotb3(pdbs$b[1,], typ="l", lwd=2, sse=pdb)
points(pdbs$b[2,], typ="l", col="red")
points(pdbs$b[3,], typ="l", col="blue")
points(pdbs$b[4,], typ="l", col="darkgreen")
points(pdbs$b[5,], typ="l", col="orange")
abline(v=100, col="gray")
```

We can improve the superposition/fitting of our models by finding the most consistent “rigid core” common across all the models. For this we will use the `core.find()` function:

```{r}
core <- core.find(pdbs)
```

We can now use the identified core atom positions as a basis for a more suitable superposition and write out the fitted structures to a directory called `corefit_structures`:

```{r}
core.inds <- print(core, vol=0.5)
```

```{r}
xyz <- pdbfit(pdbs, core.inds, outpath="corefit_structures")
```

The resulting superposed coordinates are written to a new director called `corefit_structures/`.

Now we can examine the RMSF between positions of the structure.

```{r}
rf <- rmsf(xyz)

plotb3(rf, sse=pdb)
abline(v=100, col="gray", ylab="RMSF")
```

## Predicted Alignment Error for domains

Independent of the 3D structure, AlphaFold produces an output called **Predicted Aligned Error (PAE)**.

```{r}
library(jsonlite)

pae_files <- list.files(path=results_dir,
                        pattern=".*model.*\\.json",
                        full.names = TRUE)
```

For example, reading 1st and 5th files:
```{r}
pae1 <- read_json(pae_files[1],simplifyVector = TRUE)
pae5 <- read_json(pae_files[5],simplifyVector = TRUE)

attributes(pae1)
```

```{r}
head(pae1$plddt)
```

The maximum PAE values are useful for ranking models. Here we can see that model 5 is much worse than model 1. The lower the PAE score the better. How about the other models, what are thir max PAE scores?

```{r}
pae1$max_pae
```

```{r}
pae5$max_pae
```

We can plot the N by N (where N is the number of residues) PAE scores with ggplot or with functions from the Bio3D package:

```{r}
plot.dmat(pae1$pae, 
          xlab="Residue Position (i)",
          ylab="Residue Position (j)")
```

```{r}
plot.dmat(pae5$pae, 
          xlab="Residue Position (i)",
          ylab="Residue Position (j)",
          grid.col = "black",
          zlim=c(0,30))
```

We should really plot all of these using the same z range:
```{r}
plot.dmat(pae1$pae, 
          xlab="Residue Position (i)",
          ylab="Residue Position (j)",
          grid.col = "black",
          zlim=c(0,30))
```

# Residue conservation from alignment file

```{r}
aln_file <- list.files(path=results_dir,
                       pattern=".a3m$",
                        full.names = TRUE)
aln_file
```

```{r}
aln <- read.fasta(aln_file[1], to.upper = TRUE)
```

> **Q.** How many sequences are in this alignment?

```{r}
dim(aln$ali)
```

We can score residue conservation in the alignment with the `conserv()` function.

```{r}
sim <- conserv(aln)
```

```{r}
plotb3(sim[1:99], sse=trim.pdb(pdb, chain="A"),
       ylab="Conservation Score")
```

Note the conserved Active Site residues D25, T26, G27, A28. These positions will stand out if we generate a consensus sequence with a high cutoff value:

```{r}
con <- consensus(aln, cutoff = 0.9)
con$seq
```

For a final visualization of these functionally important sites we can map this conservation score

```{r}
m1.pdb <- read.pdb(pdb_files[1])
occ <- vec2resno(c(sim[1:99], sim[1:99]), m1.pdb$atom$resno)
write.pdb(m1.pdb, o=occ, file="m1_conserv.pdb")
```
