---
title: "Class 9: Candy Mini-Project"
author: "PID: A17295169"
format: pdf
toc: true
---

## 1. Background

Today we are taking a detor to analyze a fun dataset (that we have more intrinsic insight into) with the most useful analysis method we have learned thus far - Principal Component Analysis (PCA).

## 2. Data import

The data is all related to halloween candy and is from the 538 website.

```{r}
candy_file <- "candy-data.txt"
candy <- read.csv(candy_file, row.names = 1)
head(candy)
```

> **Q1.** How many different candy types are in this dataset?

```{r}
nrow(candy)
```
There are 85 different types of candy.

> **Q2.** How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```
There are 38 fruity candy types in the dataset.

## What is your favorite candy?

One of the most interesting variables in the dataset is `winpercent`. For a given candy this value is the percentage of people who prefer this candy over another randomly chosen candy from the dataset (what 538 term a matchup). Higher values indicate a more popular candy.

For example, with Twix:
```{r}
candy["Twix", ]$winpercent
```

An alternate way is with the `dplyr` package.

```{r}
library(dplyr)

candy |> 
  filter(row.names(candy)=="Twix") |> 
  select(winpercent)
```

> **Q3.** What is your favorite candy (other than Twix) in the dataset and what is it’s winpercent value?

```{r}
library(dplyr)

candy |> 
  filter(row.names(candy)=="Sour Patch Kids") |> 
  select(winpercent)
```

My favorite candy is Sour Patch Kids, and the winpercent value is 59.864.

> **Q4.** What is the winpercent value for “Kit Kat”?

```{r}
library(dplyr)

candy |> 
  filter(row.names(candy)=="Kit Kat") |> 
  select(winpercent)
```
The winpercent value for Kit Kat is 76.7686.

> **Q5.** What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
library(dplyr)

candy |> 
  filter(row.names(candy)=="Tootsie Roll Snack Bars") |> 
  select(winpercent)
```
The winpercent value for Tootsie Roll Snack Bars is 49.6535.

**N.B:** There is a useful `skim()` function in the **skimr** package that can help give you a quick overview of a given dataset. Let’s install this package and try it on our candy data.

```{r}
library("skimr")
skim(candy)
```

A shortcut:
```{r}
skimr::skim(candy)
```

> **Q6.** Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

Yes, the column "winpercent" stands out because most of the variables (chocolate, fruity, caramel, etc.) are 0 or 1, but winpercent has values ranging from 0-100.

> **Q7.** What do you think a zero and one represent for the candy$chocolate column?

```{r}
table(candy$chocolate)
```
The meaning of 0 and 1 in `candy$chocolate` is that 1= the candy has chocolate and 0= the candy does not have chocolate.

## 3. Exploratory Analysis

> **Q8.** Plot a histogram of winpercent values using both base R and ggplot2.

```{r}
library(ggplot2)

ggplot(candy, aes(x = winpercent)) +
  geom_histogram(binwidth = 5) +
  labs(title = "Histogram of Candy Win Percent",
       x = "Win Percent",
       y = "Count")
```

> **Q9.** Is the distribution of winpercent values symmetrical?

The distrubution of winpercent values is not symmetrical, it is slightly right-skewed with the lower values.

> **Q10.** Is the center of the distribution above or below 50%?

The center of distribution is slightly above 50%.

```{r}
mean(candy$winpercent)
```

```{r}
summary(candy$winpercent)
```


> **Q11.** On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
# Line by line for chocolate candy
choc.ind <- as.logical(candy$chocolate)
choc.candy <- candy[choc.ind, ]
choc.win <- choc.candy$winpercent
mean(choc.win)
```

```{r}
# Line by line for fruity candy
fruit.ind <- as.logical(candy$fruity)
fruit.candy <- candy[fruit.ind, ]
fruit.win <- fruit.candy$winpercent
mean(fruit.win)
```

```{r}
# Both chocolate and fruity candy in 1 code
mean(candy$winpercent[as.logical(candy$chocolate)])
mean(candy$winpercent[as.logical(candy$fruity)])
```
Chocolate candy is higher on average than fruit candy:
60.92153 > 44.11974.

> **Q12.** Is this difference statistically significant?

```{r}
t.test(candy$winpercent[as.logical(candy$chocolate)],
       candy$winpercent[as.logical(candy$fruity)])
```
This difference is statistically different because the p.value 2.871e-08 is smaller than 0.05. 

## 4. Overall Candy Rankings

Let’s use the base R `order()` function together with `head()` to sort the whole dataset by a variable/column of interest. Or if you have been getting into the tidyverse and the dplyr package you can use the `arrange()` function together with `head()` to do the same thing and answer the following questions:

> **Q13.** What are the five least liked candy types in this set?

```{r}
candy |>
  arrange(winpercent) |>
  select(winpercent) |>
  head(5)
```
The five least liked candy types in this set are: 1. Nik L Nip, 2. Boston Baked Beans, 3. Chiclets, 4. Super Bubble, and 5. Jawbusters.

```{r}
x <-c(5,10,1)
order(x)
```

```{r}
x[ order (x) ]
```

```{r}
candy [ order(candy$winpercent),]
```

>**Q14.** What are the top 5 all time favorite candy types out of this set?

```{r}
candy |>
  arrange(-winpercent) |>
  head(5)
```
The five most liked candy types in this set are: 1. Reese's Peanut Butter cup, 2. Reese's Minatures, 3. Twix, 4. Kit Kat, and 5. Snickers.

> **Q15.** Make a first barplot of candy ranking based on winpercent values.

```{r,fig.height=15, fig.width=10}
ggplot(candy) +
  aes(winpercent, rownames(candy)) +
  geom_col()
```

> **Q16.** This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r,fig.height=15, fig.width=10}
ggplot(candy) +
  aes(winpercent, reorder(rownames(candy), winpercent)) +
  geom_col()
```

## Adding useful color

We start by making a vector of all black values (one for each candy). Then we overwrite chocolate (for chocolate candy), brown (for candy bars) and red (for fruity candy) values.

```{r}
my_cols=rep("black", nrow(candy))

## Chocolate candy
my_cols[as.logical(candy$chocolate)] <- "chocolate"
## Candy bars
my_cols[as.logical(candy$bar)] <- "brown"
## Fruity candy
my_cols[as.logical(candy$fruity)] <- "red"
```

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) +
  ylab("")
```

> **Q17.** What is the worst ranked chocolate candy?

```{r}
choc <- candy[as.logical(candy$chocolate), ]
choc[which.min(choc$winpercent), ]
```
The worst ranked chocolate candy is Sixlets.

> **Q18.** What is the best ranked fruity candy?

```{r}
fruity <- candy[as.logical(candy$fruity), ]
fruity[which.max(fruity$winpercent), ]
```
The best ranked fruity candy is Starburst.

## 5. Taking a look at pricepercent

Make a plot of winpercent vs pricepercent

```{r}
library(ggrepel)

ggplot(candy) +
  aes(x = pricepercent, y = winpercent, label = rownames(candy)) +
  geom_point(col = my_cols) + 
  geom_text_repel(col = my_cols, size = 3.3, max.overlaps = 5)
```

> **Q19.** Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

```{r}
candy[which.max(candy$winpercent / candy$pricepercent), ]
```
Tootsie Roll Snack Bars is the highest ranked in terms of winpercent for the least money.

> **Q20.** What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```

```{r}
library(dplyr)
candy |>
  arrange(-pricepercent) |> 
  select(pricepercent, winpercent) |> 
  head(n=5)
```
The top 5 most expensive candy types are: 1. Nik L Nips, 2. Nestle Smarties, 3. Ring pops, 4. Hershey's Krackle's, and 5. Hershey's Milk Chocolates. Among all of these, the least popular is Nik L Nips because it has the lowest winpercent (22.44534).

## 6. Exploring the correlation structure

We can calculate the pair-wise correlation of all our columns

> **N.B:**  This matters especially because PCA operates not on raw data but on the relationships between variables. Specifically, it finds the directions that capture the most co-variation in your data. This is also why we will use `scale=TRUE` when calling `prcomp()` in the next section: it ensures all variables contribute equally by working from correlations (standardized and scale-free) rather than raw covariances, which would be dominated by whichever variable happens to have the largest numeric range. Can you guess which variable in our dataset that would be?
An important point here is to think of the analysis below as a preview of your upcoming PCA results: the patterns we spot here should reappear in the loadings plot at the end of the next section.

```{r}
library(corrplot)
cij <- cor(candy)
corrplot(cij)
```

A few things to notice about this correlation matrix plot. First, the diagonal is always perfectly blue (correlation = 1.0) — every variable is perfectly correlated with itself. Second, the matrix is symmetric: the correlation between `chocolate` and `bar` is the same as between `bar` and `chocolate`, so the upper and lower triangles are mirror images. Most importantly, this matrix is exactly what PCA will decompose mathematically. When we run `prcomp()` on scaled data, it is performing an eigen-decomposition of this correlation matrix — finding the directions (eigen-vectors) that explain the most variance (eigen-values). The loading values you will see shortly are those eigen-vectors, and the proportion of variance explained by each PC corresponds to its eigen-value. The corrplot is therefore not just an exploratory tool, it is the structure that PCA will summarize.

> **Q22.** Examining this plot what two variables are anti-correlated (i.e. have minus values)?

The two variables that are anti-correlated are chocolate and fruity. This occurs because candies are either chocolatey or fruity.

> **Q23.** Use your corrplot result to make a prediction: which variables do you expect will have the largest contributions (i.e. loadings) to PC1 (i.e., drive the most separation between candies along the first principal component)?

Variables that would have the largest contributions to PC1: chocolate and bar (positively) and fruity (negatively).

## Principal Component Analysis

In this case we want to be sure to set `scale=TRUE` argument for `prcomp()` because we have one variable `winpercent` that is on a very different scale than all others and would otherwise dominate our PCA results.

```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```

First major result figure is the "score plt" of PC1 vs PC2
- how the different candy are related to each other on our new PC axis:

```{r}
pca$x
```

```{r}
ggplot(pca$x) +
  aes(PC1, PC2, label=row.names(pca$x)) +
  geom_point(col=my_cols) 
```

We can make a much nicer plot with the ggplot2 package.

```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])
```

```{r}
p <- ggplot(my_data) + 
  aes(x = PC1, y = PC2, 
      size = winpercent/100,  
      text = rownames(my_data),
      label = rownames(my_data)) +
  geom_point(col = my_cols)

p
```

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

The second major results figure from PCA is the so-called "loadings plot.

```{r}
ggplot(pca$rotation) +
  aes(PC1, 
      reorder(row.names(pca$rotation), PC1)) +
  geom_col()
```

> **Q24.** Complete the code to generate the loadings plot above. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you? Where did you see this relationship highlighted previously?

Chocolate and bar are picked up strongly as a positive contributer and fruity is picked up strongly as a negative contributor by PC1 in the positive direction. This makes sense because the same pattern is seen in the correlation plot.

> **Q25.** Based on your exploratory analysis, correlation findings, and PCA results, what combination of characteristics appears to make a “winning” candy? How do these different analyses (visualization, correlation, PCA) support or complement each other in reaching this conclusion?

"Winning" candies tend to have chocolate and are candy bars, but are not fruity candies. Additionally, they have decent pricings. The different analyses support this through visualization of the histograms and bar plots showing a higher win percent for chocolate candies. Moreover, a t-test was performed which validated that chocolate candies are significantly more popular. The correlation analysis showed chocolate being the candy that was positively associated with popularity, and PCA showed PC1 separating chocolate vs fruity candies with chocolate being the more popular candy. Hence, chocolate candies is more preferred and are the "winning" candies.
