---
title: "Class 12 Lab Session"
author: "Mikaela Joya (PID: A17295169)"
format: pdf
toc: true
---

# Section 1. Proportion of G/G in a population

Downloaded a CSV file from Ensemble < https://www.ensembl.org/Homo_sapiens/Variation/Sample?db=core;r=17:39894595-39895595;v=rs8067378;vdb=variation;vf=959672880#373531_tablePanel >

Here we read this CSV file

```{r}
mxl <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(mxl)
```
```{r}
table(mxl$Genotype..forward.strand.)
```

```{r}
table(mxl$Genotype..forward.strand.) / nrow(mxl) * 100
```

Now let's look at a different population. I picked the GBR.

```{r}
gbr <- read.csv("373522-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(gbr)
```

Find proportion of G|G

```{r}
round(table(gbr$Genotype..forward.strand.) /nrow(gbr) * 100, 2 )
```

This variant that is associated with childhood asthma is more frequent in the GBR population than the MKL population.

Let's now dig into this further.

# Section 4: Population Scale Analysis

> **Q13.** Read this file into R and determine the sample size for each genotype and their
corresponding median expression levels for each of these genotypes.

```{r}
file <- "rs8067378_ENSG00000172057.6.txt"
data <- read.table(file)

colnames(data) <- c("sample", "genotype", "expression")

head(data)
```

```{r}
table(data$genotype)
```

The sample size for the A/A genotype is 108, the A/G genotype is 233, and G/G genotype is 121.

```{r}
tapply(data$expression, data$genotype, median)
```

The corresponding median expressions are 31.24847 (A/A), 25.06486 (A/G), and 20.07363 (G/G).

> **Q14.**  Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

```{r}
library(ggplot2)

ggplot(data,
       aes(x = genotype, y = expression, fill = genotype)) + 
       geom_boxplot(notch = TRUE) + 
       geom_jitter(width = 0.15, alpha = 0.5) +
  labs(
    title = "ORMDL3 Expression by Genotype",
    x = "Genotype",
    y = "Expression") +
  theme_gray()
```
This boxplot shows that the A/A genotype individuals have higher expression levels of ORMDL3 compared to those with G/G. This means that the rs8067378 SNP affects the expression of the ORMDL3 gene because the G allele is associated with a decrease in gene expression.