---
title: "Class 5: Data Viz with ggplot2"
author: "Mikaela Joya (PID: A17295169)"
format: pdf
toc: true
---

## Background

There are many graphics systems in R for making plots and figures. These include so-called *"base R" graphics* like the `plot()` function and add on packages like **ggplot2**.

Let's compare how we make a simple figure with these two systems:

We can use the in-built `cars` dataset:

```{r}
head ( cars )
```

```{r}
plot(cars)
```

Before I can use ggplot 2 I need to install it on my computer.
To do this we can use the function `install.packages("ggplot2")`

> **N.B.** We never run `install.packages()` in our quarto doc (we run it once only in our R console) as it would re-install the package everytime we render our quarto report.

Once installed we need to load up the package in our R brain:
```{r}
library(ggplot2)
```

The main function in the **ggplot2** package is called `ggplot()`






Every ggplot has at least 3 layers:

- the **data** (a data.frame of the stuff we want to plot)
- the **aes**thetics (how the data maps to the plot)
- the **geom** layer (how you want the plot drawn, e.g. points, lines, etc.)

```{r}
ggplot(cars) +
  aes(x=speed , y=dist ) +
  geom_point()
```

## Add some custom features

Let's add a trend line that shows the relationship between speed and distance.

```{r}
ggplot(cars) +
  aes(x = speed , y = dist ) +
  geom_point() +
  geom_smooth() +
  theme_bw() +
  labs(title = "Stopping Distance of Old Cars",
    x = "Speed (MPH)",
    y = "Distance (ft)")
```

Q. Can you make the `geom_smooth()` function produce a linear straight line fit to the data and turn off the "gray" error region.

```{r}
ggplot(cars) +
  aes(x = speed , y = dist ) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  theme_bw() +
  labs(title = "Stopping Distance of Old Cars",
    x = "Speed (MPH)",
    y = "Distance (ft)")
```

---

## Gene expression figure

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

```{r}
sum(genes$State == "up")
```

A useful new function in the `table()` function:

```{r}
table(genes$State)
```

My first plot attempt
```{r}
ggplot(genes) +
  aes(Condition1, Condition2, col=State) +
  geom_point() +
  scale_color_manual(values=c("purple","blue","pink")) +
  theme_bw() +
  labs(title="Expression changes upon GLP-1 inhibitor treatment",
         x="Control (no drug) ",
         y="Drug Treatment")
```

## Going further

Here we read the famous gapmider dataset:

```{r}
# File location online
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
head(gapminder)
```

> Q. How many entries (i.e. rows) are in this dataset?

```{r}
nrow(gapminder)
```

> Q. How many different "country" entries are in this dataset?

```{r}
length(unique(gapminder$country
```
```{r}
length( unique(gapminder$country))
```

Let's make our first plot of the entire dataset:

Plot of "gdpPercap" vs "lifeExp" colored by "continent"

```{r}
p <- ggplot(gapminder) +
  aes(x=gdpPercap, y=lifeExp, col=continent) +
  geom_point(alpha=0.3)
```

```{r}
p
```

I can add more layers to `p`
```{r}
p +
  facet_wrap(~continent)
```

Make a plot for 1977 and 2007 only (not all the years in the dataset).

> Q. First use the **dplyr** package and the `filter()` function from that package to extract the year 2007.

```{r, message=FALSE, warning=FALSE}
library(dplyr)
```

```{r}
g07 <- filter(gapminder, year == 2007)
g77 <- filter(gapminder, year == 1977)
g <- filter(gapminder, year == 2007 | year == 1977)
```

```{r}
ggplot(g) +
  aes(gdpPercap, lifeExp, col=continent, size=pop) +
  geom_point() +
  facet_wrap(~year)
```

> Q. Make a histogram of lifeExp colored by continent (try using `fill=continent` or `col=continent`)
> Q. Make a histogram of lifeExp faceted by continent

```{r}
ggplot(gapminder) +
  aes(lifeExp, fill=continent) +
  geom_histogram() +
  facet_wrap(~continent)
```
