---
title: "HW Class 6 Q6 (R Functions)" 
author: "Mikaela Joya (PID: A17295169)"
format: pdf
toc: TRUE
---

# **Q6.** How would you generalize the original code above to work with any set of input protein structures?

The code could be generalized by writing a function that removes the repeated steps for each protein. This can be done by taking a PDB ID as an input. This function allows the PDB file to be read, trims the structure to the selected chain and atoms, extracts the B-factor values, and generates a plot for the specified proteins. Additionally, returning the B-factor vector allows the results to be used for more comparison or clustering.

## Write your own function starting from the code above that analyzes protein drug interactions by reading in any protein PDB data and outputs a plot for the specified protein.

The code given is generalized below:

```{r}
library(bio3d)
```

```{r}
# analyze_protein() 
# Input: 
# pdb_id is a character string that represents PDB ID, 
# chain_id is the protein chain used to analyze (default= "A"),
# atom_type is the atom type to select (default= "CA")
# Function: 
# Reads the protein structure from the PDB database, 
# trims to a specified chain and atom type,
# Then it extracts the B-factor values and plots the B-factors
# Output: Numeric vector with B-factor values for the selected atoms
analyze_protein <- function(pdb_id, chain_id = "A", atom_type = "CA") {
  pdb <- read.pdb(pdb_id)
  pdb_trim <- trim.pdb(pdb, chain = chain_id, elety = atom_type)
  b_factors <- pdb_trim$atom$b
  
  plotb3(b_factors, sse= pdb_trim, typ = "l", ylab = "Bfactor")
  
  return(b_factors)
}
```

```{r}
# Example usage
s1.b <- analyze_protein("4AKE") # kinase with drug
s2.b <- analyze_protein("1AKE") # kinase no drug
s3.b <- analyze_protein("1E4Y") # kinase with drug
```

